---
title: "cvxDesign: Optimal experimental designs via convex optimization"
date: last-modified
format:
  gfm:
    toc: false
execute:
  echo: true
  warning: false
  message: false
---

<!-- badges: start -->
<!-- [![R-CMD-check](https://github.com/chikuang/cvxDesign/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/chikuang/cvxDesign/actions/workflows/R-CMD-check.yaml) -->
[![](https://img.shields.io/github/languages/code-size/chikuang/cvxDesign.svg)](https://github.com/chikuang/cvxDesign)
<!-- badges: end -->

**Authors:**  
Chi-Kuang Yeh (Georgia State University)  
  [![ORCID](https://img.shields.io/badge/ORCID-0000--0001--7057--2096-A6CE39?logo=orcid)](https://orcid.org/0000-0001-7057-2096)


Julie Zhou (University of Victoria)



``` {r}
# options(scipen = 999)
options(digits = 4)
```

## Description

`cvxDesign` is an R package for computing optimal experimental designs using convex optimization.  
The package is intended for researchers and practitioners working on approximate and exact design problems under a variety of criteria, including classical criteria such as D-, A-, E-, and c-optimality, as well as compound and multi-objective design problems.

The package is motivated by modern design settings where efficient numerical optimization, equivalence-theorem-based verification, and flexible model specification are all essential.

## Installation

You can install the development version of `cvxDesign` from GitHub by running

```r
# install.packages("pak")
pak::pak("chikuang/cvxDesign")
```

or

```r
# install.packages("remotes")
remotes::install_github("chikuang/cvxDesign")
```

## Details

Consider a regression model of the form

$$
y_i = \eta(\mathbf{x}_i, \boldsymbol{\theta}) + \varepsilon_i, \quad i = 1, \ldots, n,
$$

where $y_i$ is the response observed at design point $\mathbf{x}_i \in \mathcal{X}$,  
$\boldsymbol{\theta} \in \mathbb{R}^p$ is an unknown parameter vector, and $\eta(\mathbf{x}, \boldsymbol{\theta})$ is the mean response function, which may be linear or nonlinear in the parameters.

Optimal design seeks a design measure $\xi$ over the design space $\mathcal{X}$ that optimizes a scalar criterion based on the information matrix $M(\xi)$. Common criteria include:

- **D-optimality**, which maximizes $\log \det M(\xi)$;
- **A-optimality**, which minimizes $\mathrm{tr}\{M^{-1}(\xi)\}$;
- **E-optimality**, which maximizes the minimum eigenvalue of $M(\xi)$;
- **c-optimality**, which minimizes the variance of estimating a specified linear combination of parameters.

In many applications, the design problem can be formulated and solved using convex optimization. This makes it possible to handle a wide range of criteria and constraints in a unified computational framework.

The goal of `cvxDesign` is to provide a clean and extensible interface for these computations, together with tools for:

- computing optimal approximate designs,
- studying exact designs on finite candidate sets,
- evaluating design efficiency,
- checking equivalence theorems,
- visualizing directional derivative or sensitivity functions.

## Examples

First load the package.

``` {r}
# load package
library(cvxDesign)
```

### D-optimal design for a quadratic regression model

Consider the quadratic regression model

$$
y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \varepsilon_i.
$$

We first define the regression function.

```{r}
quad_reg <- function(x) {
  c(1, x, x^2)
}
```

Suppose the candidate design space is a finite grid on $[-1,1]$.

```{r}
u <- seq(-1, 1, length.out = 101)
head(u)
```

A typical workflow in `cvxDesign` will look like the following.

```{r Dopt}
dout <- calc_Dopt(
  u = u,
  f = quad_reg,
  drop_tol = 1e-4
)

dout$design |> round(3)
dout$value
dout$status
```

The returned object is expected to contain at least:

- the selected support points,
- the corresponding optimal weights,
- the value of the design criterion,
- and, when requested, information useful for equivalence-theorem checks.

### A-optimal design

Similarly, one may compute the A-optimal design by changing the criterion.

```{r Aopt}
aout <- calc_Aopt(
  u = u,
  f = quad_reg,
  drop_tol = 1e-4
)

aout$design |> round(3)
aout$value
aout$status
```

### c-optimal design

If the target is a specific linear combination $c^\top \theta$, then a c-optimal design can be computed by supplying the vector `cvec`.

```{r copt}
cout <- calc_copt(
  u = u,
  f = quad_reg,
  cVec = c(0.3, 0.4, 0.3),
  drop_tol = 1e-4
)

cout$design |> round(3)
cout$value
cout$status
```

### Equivalence theorem

An important part of optimal design computation is verification.  
The package is designed to support equivalence-theorem-based diagnostics and plotting.

``` {r equivalence for A- and D-optimalities}
quad_reg <- function(x) c(1, x, x^2)
u <- seq(-1, 1, length.out = 101)

dout <- calc_Dopt(u, quad_reg, drop_tol = 1e-4)
eq_d <- check_equivalence(dout, f = quad_reg)

print(eq_d)
plot_equivalence(eq_d)

# Aopt
aout <- calc_Aopt(u, quad_reg, drop_tol = 1e-4)
eq_a <- check_equivalence(aout, f = quad_reg)
plot_equivalence(eq_a)
```

``` {r copt equivalence}

# copt
cout <- calc_copt(u, quad_reg, cVec = c(1, 0.5, 0),
                  drop_tol = 1e-6)
eq_c <- check_equivalence(cout, f = quad_reg)
plot_equivalence(eq_c)
```

## For non-linear models

For non-linear models, the information matrix depends on the unknown parameters. 
In this case, the package supports local optimal design computation at a specified nominal parameter value. For instance, the peleg model is given as 
$$
y_i = \frac{\theta_1 x_i}{\theta_2 + x_i } + \varepsilon_i.
$$

One needs to calculate the partial derivatives of the mean function with respect to the parameters, which will be used in the information matrix calculation. In this case, it is 
$$
\frac{\partial \eta}{\partial \theta_1} = \frac{x}{\theta_2 + x}, \quad
\frac{\partial \eta}{\partial \theta_2} = -\frac{\theta_1 x}{(\theta_2 + x)^2}.
$$

```{r peleg Dopt}
peleg_reg <- function(x, theta) {
    d1 <- x / (theta[2] + x)
    d2 <- -theta[1] * x / (theta[2] + x)^2
    c(d1, d2)
}
u <- seq(0.1, 100, length.out = 1001)
theta_nom <- c(0.5, 0.05)
dout <- calc_Dopt(u, 
                  function(x) peleg_reg(x, theta = theta_nom), 
                  drop_tol = 1e-4)
dout$design |>
  dplyr::filter(.data$weight > 1e-3) |>
  round(3)
dout$value
dout$status
```

## For multiple objective design

### Maximin design

The maximin design is a robust design that maximizes the minimum efficiency across a set of candidate parameter values or different design criteria. For instance, one may want to find a design that performs well across multiple nominal parameter values in a non-linear model. This can be achieved by specifying a set of candidate parameter values and using the `calc_maximin` function.

```{r maximin Dopt}
library(cvxDesign)

## regression function
quad_reg <- function(x) c(1, x, x^2)

## candidate set
u <- seq(-1, 1, length.out = 41)
```


### Step 1: single-objective designs


``` {r}

res_D <- compute_design_SO(
  u = u,
  f = quad_reg,
  criterion = "D"
)

res_A <- compute_design_SO(
  u = u,
  f = quad_reg,
  criterion = "A"
)

## reference losses
loss_ref <- list(
  D = res_D$loss,
  A = res_A$loss
)

print(loss_ref)
```

### Step 2: maximin design

``` {r}

res_DA <- compute_maximin_design(
  u = u,
  f = quad_reg,
  loss_ref = loss_ref,
  criteria = c("D", "A")
)
```

### Step 3: inspect result

``` {r}

print(res_DA$design)
print(res_DA$loss)
print(res_DA$efficiency)

cat("tstar =", res_DA$tstar, "\n")
cat("1 / tstar =", 1 / res_DA$tstar, "\n")
cat("min efficiency =", min(res_DA$efficiency), "\n")
```


### Step 4: numerical checks

``` {r}

tol <- 1e-4

eq1 <- abs(res_DA$efficiency["D"] - res_DA$efficiency["A"])
eq2 <- abs(min(res_DA$efficiency) - 1 / res_DA$tstar)

cat("|eff_D - eff_A| =", eq1, "\n")
cat("|min(eff) - 1/tstar| =", eq2, "\n")

stopifnot(eq1 < tol)
stopifnot(eq2 < tol)

cat("All checks passed.\n")
```

### Equivalence theorem


``` {r multi-equivalence theorem}
## =========================================================
## Maximin multi-criterion design: D + A + c
## =========================================================

quad_reg <- function(x) c(1, x, x^2)
u <- seq(-1, 1, length.out = 41)

## target contrast for c-optimality
cvec <- c(0, 1, 0)

## ---------------------------------------------------------
## Step 1: single-objective reference designs
## ---------------------------------------------------------

res_D <- compute_design_SO(
  u = u,
  f = quad_reg,
  criterion = "D"
)

res_A <- compute_design_SO(
  u = u,
  f = quad_reg,
  criterion = "A"
)

res_c <- compute_design_SO(
  u = u,
  f = quad_reg,
  criterion = "c",
  opts = list(cVec_c = cvec)
)

## reference losses
loss_ref <- list(
  D = res_D$loss,
  A = res_A$loss,
  c = res_c$loss
)

print(loss_ref)

## ---------------------------------------------------------
## Step 2: maximin design
## ---------------------------------------------------------

res_DAc <- compute_maximin_design(
  u = u,
  f = quad_reg,
  loss_ref = loss_ref,
  criteria = c("D", "A", "c"),
  opts = list(cVec_c = cvec)
)

print(res_DAc$design)
print(res_DAc$loss)
print(res_DAc$efficiency)

cat("tstar =", res_DAc$tstar, "\n")
cat("1 / tstar =", 1 / res_DAc$tstar, "\n")
cat("min efficiency =", min(res_DAc$efficiency), "\n")
```

### Step 3: directional derivatives

``` {r step3}

dd_DAc <- calc_directional_derivatives(
  u = u,
  M = res_DAc$info_matrix,
  f = quad_reg,
  criteria = c("D", "A", "c"),
  cVec = cvec
)

```

### Step 4: eta weights for maximin equivalence theorem

``` {r step4}

eta_DAc <- calc_eta_weights_maximin(
  tstar = res_DAc$tstar,
  loss_ref = loss_ref,
  loss_model = res_DAc$loss,
  directional_derivatives = dd_DAc,
  criteria = c("D", "A", "c"),
  q = length(quad_reg(0))
)

print(eta_DAc)
```

### Step 5: equivalence theorem check

``` {r step5}

eq_DAc <- check_equivalence_maximin(
  design_obj = res_DAc,
  directional_derivatives = dd_DAc,
  eta = eta_DAc
)

print(eq_DAc)
```

### Step 6: plot

``` {r step 6}

plot_equivalence_maximin(
  design_obj = res_DAc,
  directional_derivatives = dd_DAc,
  eta = eta_DAc,
  criteria = c("D", "A", "c"),
  cex_lab = 1.5,
  cex_axis = 1.3,
  cex_main = 1.2,
  mar = c(5, 7, 3, 1)
)
```

## Planned features

- [x] Basic package infrastructure
- [x] Core support for convex-optimization-based design computation
- [x] D-optimality
- [x] A-optimality
- [ ] E-optimality
- [x] c-optimality
- [x] Compound and multi-objective criteria
- [x] Equivalence theorem (sensitivity functions) diagnostics
- [ ] Exact design support on finite candidate sets
- [ ] Benchmark examples and vignettes
- [ ] Shiny app
- [ ] JSS paper companion materials

## Package vision

The long-term goal of `cvxDesign` is to provide a general computational framework for modern optimal design problems, including:

- classical regression design,
- heteroscedastic models,
- constrained design problems,
- compound and robust criteria,
- group testing design,
- and other application-driven problems arising in statistics, biostatistics, and machine learning.

## References

1. Atkinson, Anthony C., Donev, Aleksandar N., and Tobias, Randall D. (2007). *Optimum Experimental Designs, with SAS*. Oxford University Press.
2. Fedorov, Valerii V. (1972). *Theory of Optimal Experiments*. Academic Press.
3. Pukelsheim, Friedrich. (2006). *Optimal Design of Experiments*. SIAM.
4. Silvey, Samuel D. (1980). *Optimal Design*. Chapman and Hall.
5. Yeh, Chi-Kuang and collaborators. Ongoing work on convex-optimization-based methods for compound and application-driven optimal design problems.

